\textbf{ Interpretable Gaussian Processes for Stellar Light Curves }

### 2020-12-03 tex/gitlinks.tex
revert eps3, rename to eps

\newcommand{\codelink}[1]{\href{https://github.com/rodluger/starry_process/blob/ae47501186fb2c08dfe0ce5e5d8355fab2d9bfad/tex/figures/#1.py}{\codeicon}\,\,}
\newcommand{\animlink}[1]{\href{https://github.com/rodluger/starry_process/blob/ae47501186fb2c08dfe0ce5e5d8355fab2d9bfad/tex/figures/#1.gif}{\animicon}\,\,}
\newcommand{\prooflink}[1]{\href{https://github.com/rodluger/starry_process/blob/ae47501186fb2c08dfe0ce5e5d8355fab2d9bfad/tex/tests/#1.py}{\raisebox{-0.1em}{\input{tests/#1.tex}}}}
\newcommand{\cilink}[1]{\href{https://dev.azure.com/rodluger/starry_process/_build}{#1}}

>>>>

\newcommand{\codelink}[1]{\href{https://github.com/rodluger/starry_process/blob/56f89997cccb24bacd2c3b8d1b6d990961902783/tex/figures/#1.py}{\codeicon}\,\,}
\newcommand{\animlink}[1]{\href{https://github.com/rodluger/starry_process/blob/56f89997cccb24bacd2c3b8d1b6d990961902783/tex/figures/#1.gif}{\animicon}\,\,}
\newcommand{\prooflink}[1]{\href{https://github.com/rodluger/starry_process/blob/56f89997cccb24bacd2c3b8d1b6d990961902783/tex/tests/#1.py}{\raisebox{-0.1em}{\input{tests/#1.tex}}}}
\newcommand{\cilink}[1]{\href{https://dev.azure.com/rodluger/starry_process/_build}{#1}}

### 2020-12-03 tex/ms.tex
revert eps3, rename to eps

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble.pdf}
        \oscaption{nullspace_ensemble}{%
            Posterior shrinkage as a function of spherical harmonic degree
            for a hypothetical scenario in which an observer can measure a
            stellar light curve at a single inclination (thin blue curves),
            at 3 different inclinations (thin orange curves), and at 10 different
            inclinations (thin green curves). The thicker curves correspond to
            the mean shrinkage for each experiment. The inclinations are
            drawn randomly from an isotropic distribution.
            The information content in the light curve of a star observed
            from a single vantage point approaches zero as $l$
            increases. However, observing the star from many vantage points
            allows one to recover nearly all of the information in the
            even spherical harmonic modes.
            While this thought experiment is not possible in practice for
            a single star, it is extremely relevant to ensemble analyses of statistically
            similar stars.
            \label{fig:nullspace_ensemble}
        }
    \end{centering}
\end{figure}

@@

(Appendix \ref{sec:lat}).

@@

\subsection{Marginalizing over inclination}
\label{sec:inclination}
%
As we mentioned in the previous section,
the equations for the mean and covariance of the flux GP
(Equations~\ref{eq:mean_f} and \ref{eq:cov_f}, respectively) are conditioned
on specific values of the stellar properties $\pmb{\theta}_\star$ and the
spot properties $\pmb{\theta}_\bullet$. To obtain the posterior distribution
for these parameters, we must typically resort to numerical sampling techniques,
which often scale steeply with the number of parameters. It is therefore generally
desirable to keep the total number of parameters small, especially when
employing the GP in an ensemble setting.
In such a setting, we might have light curves from $N_\star$ stars, all of which
we believe to have similar spot properties (perhaps because they have
similar spectral types and rotation periods, for example).
The total number of parameters in our problem is therefore
%
\begin{align}
    N_\theta = N_\star \, \mathrm{dim}(\pmb{\theta}_\star) + 5
    \quad,
\end{align}
%
since each of the stars will have their own set of stellar properties
$\pmb{\theta}_\star$ but will all share the same 5 spot properties
$\pmb{\theta}_\bullet$ (by assumption).
In a typical problem, the stellar parameters might be
%
\begin{align}
    \pmb{\theta}_\star =
    \left(
    i \,\,\,
    P \,\,\,
    u_1 \,\,\,
    u_2
    \right)^\top
    \quad,
\end{align}
%
i.e., the inclination, the rotational period, and some (usually two) limb
darkening coefficients. Thus, for a reasonably sized ensemble of $N_\star=100$
stars, we would have to sample over $N_\theta = 405$ parameters.
While large, this number is certainly not absurd, especially by modern standards.
However, it does pose
a problem when considering how complex the posterior distribution for the
spot mapping problem can be. In addition to strong nonlinear degeneracies
between some of the parameters (such as the contrast $c$ and the
number of spots $N$), the posterior is often multimodal, especially in the
stellar inclinations. While modern sampling schemes such as Hamiltonian
Monte Carlo and Nested Sampling may in principle be able to deal with these
issues, in practice it can be very difficult to obtain convergence in a
reasonable amount of time.

@@

Fortunately, there is a better way to reduce the number of parameters in
the problem: we can explicitly marginalize over the stellar inclination.
That is, we can write the first two moments of our flux GP as
%
\begin{align}
    \label{eq:mu_marg}
    \mathrm{E} \Big[ \mathbf{f} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
     & =
    \int
    \pmb{\mathcal{A}}(i, \bar{\pmb{\theta}}_\star) \,
    \mathrm{E} \Big[ \mathbf{y} \, \Big| \, \pmb{\theta}_\bullet \Big]
    p(i) \, \mathrm{d}i \\
    \label{eq:cov_marg}
    \mathrm{E} \Big[ \mathbf{f} \, \mathbf{f}^\top \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
    - \mathbf{1} \mathbf{1}^\top
     & =
    \int
    \pmb{\mathcal{A}}(i, \bar{\pmb{\theta}}_\star) \,
    \mathrm{E} \Big[ \mathbf{y} \, \mathbf{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big] \,
    \pmb{\mathcal{A}}^\top(i, \bar{\pmb{\theta}}_\star)
    p(i) \, \mathrm{d}i
    \quad,
\end{align}
%
where the expectations inside the integrals are given by
Equations~(\ref{eq:exp_y_sep}) and (\ref{eq:exp_yy_sep}), respectively, and
$\bar{\pmb{\theta}}_\star$ are all the stellar parameters \emph{except}
the inclination, i.e.,
%
\begin{align}
    \bar{\pmb{\theta}}_\star = \left(
    P \,\,\,
    u_1 \,\,\,
    u_2
    \right)^\top
    \quad,
\end{align}
%
where perhaps $P$ is fixed for each star and the limb darkening coefficients
are shared among all stars, as described above.
%
Thus, if we are able to perform the integrals in Equations~(\ref{eq:mu_marg})
and (\ref{eq:cov_marg}), we can \emph{dramatically} reduce the number of
parameters in our ensemble problem.
%
As we show in Appendix~\ref{sec:inclination}, if we assume that stellar
inclinations are distributed isotropically, these integrals
do in fact have closed-form solutions,
since they are (again) integrals over Wigner matrices.
%
We may then replace our expressions for the GP mean and covariance with
%
\begin{align}
    \pmb{\mu}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbf{f} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
    \\
    \pmb{\Sigma}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbf{f} \, \mathbf{f}^\top \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big] - \pmb{\mu}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet) \pmb{\mu}^\top(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
    \quad.
\end{align}
%
and use these to compute the marginal likelihood instead.

@@

\section{Computing the flux}
\label{sec:starry}
\xxx{Coming soon.}

@@

\xxx{...}

Although we adopt $a$ and $b$ as the actual inputs to our Gaussian process
(i.e., as elements of the input vector $\pmb{\theta}_\bullet$), these
parameters are not themselves ``interpretable'', as they do not directly
correspond to physical
quantities of interest such as the mean and standard deviation of the spot
latitude distribution. However, over most of the parameter space in $a$ and $b$,
the spot latitude distribution is well approximated
by a bimodal Gaussian. In particular, there exists
a one-to-one relationship between $a$ and $b$ and the mean $\mu_\phi$
and standard deviation $\sigma_\phi$ of a normal approximation to the
distribution. Moreover, we find that
if we let $\mu_\phi$ be the \emph{mode} of the PDF and
$\sigma_\phi^2$ be a \emph{local} approximation to the variance of the PDF,
the relationship has a convenient closed form.

@@

\begin{align}
    p(i) = \sin i
    \quad.
\end{align}

\subsection{First moment}
\xxx{Coming soon.}

>>>>

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble.pdf}
        \oscaption{nullspace_ensemble}{%
            Posterior shrinkage as a function of spherical harmonic degree
            for a hypothetical scenario in which an observer can measure a
            stellar light curve at a single random inclination (thin blue curves),
            at 3 random inclinations (thin orange curves), and at 10 random
            inclinations (thin green curves). The thicker curves correspond to
            the mean shrinkage for each experiment.
            The information content in the light curve of a star observed
            from a single vantage point approaches zero as $l$
            increases. However, observing the star from many vantage points
            allows one to recover nearly all of the information in the
            even spherical harmonic modes.
            While this thought experiment is not possible in practice for
            a single star, it is extremely relevant to ensemble analyses of statistically
            similar stars.
            \label{fig:nullspace_ensemble}
        }
    \end{centering}
\end{figure}

@@

(Appendix \ref{sec:lat-transform}).

@@

\subsection{Marginalizing over inclination}
\label{sec:inclination}
%
As we mentioned in the previous section,
the equations for the mean and covariance of the flux GP
(Equations~\ref{eq:mean_f} and \ref{eq:cov_f}, respectively) are conditioned
on specific values of the stellar properties $\pmb{\theta}_\star$ and the
spot properties $\pmb{\theta}_\bullet$. To obtain the posterior distribution
for these parameters, we must typically resort to numerical sampling techniques,
which often scale steeply with the number of parameters. It is therefore generally
desirable to keep the total number of parameters small, especially when
employing the GP in an ensemble setting.
In such a setting, we might have light curves from $N_\star$ stars, all of which
we believe to have similar spot properties (perhaps because they have
similar spectral types and rotation periods, for example).
The total number of parameters in our problem is therefore
%
\begin{align}
    N_\theta = N_\star \, \mathrm{dim}(\pmb{\theta}_\star) + 5
    \quad,
\end{align}
%
since each of the stars will have their own set of stellar properties
$\pmb{\theta}_\star$ but will all share the same 5 spot properties
$\pmb{\theta}_\bullet$ (by assumption).
In a typical problem, the stellar parameters might be
%
\begin{align}
    \pmb{\theta}_\star =
    \left(
    i \,\,\,
    P \,\,\,
    u_1 \,\,\,
    u_2
    \right)^\top
    \quad,
\end{align}
%
i.e., the inclination, the rotational period, and some (usually two) limb
darkening coefficients. Thus, for a reasonably sized ensemble of $N_\star=100$
stars, we would have to sample over $N_\theta = 405$ parameters.
While large, this number is certainly not absurd, especially by modern standards.
However, it does pose
a problem when considering how complex the posterior distribution for the
spot mapping problem can be. In addition to strong nonlinear degeneracies
between some of the parameters (such as the contrast $c$ and the
number of spots $N$), the posterior is often multimodal, especially in the
stellar inclinations. While modern sampling schemes such as Hamiltonian
M0onte Carlo and Nested Sampling may in principle be able to deal with these
issues, in practice it can be very difficult to obtain convergence in a
reasonable amount of time.

@@

Fortunately, there is a better way to reduce the number of parameters in
the problem: we can explicitly marginalize over the stellar inclination.
That is, we can write the first two moments of our flux GP as
%
\begin{align}
    \label{eq:einc}
    \mathrm{E} \Big[ \mathbf{f} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
     & =
    \int
    \pmb{\mathcal{A}}(i, \bar{\pmb{\theta}}_\star) \,
    \mathrm{E} \Big[ \mathbf{y} \, \Big| \, \pmb{\theta}_\bullet \Big]
    p(i) \, \mathrm{d}i \\
    \label{eq:Einc}
    \mathrm{E} \Big[ \mathbf{f} \, \mathbf{f}^\top \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
     & =
    \int
    \pmb{\mathcal{A}}(i, \bar{\pmb{\theta}}_\star) \,
    \mathrm{E} \Big[ \mathbf{y} \, \mathbf{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big] \,
    \pmb{\mathcal{A}}^\top(i, \bar{\pmb{\theta}}_\star)
    p(i) \, \mathrm{d}i
    \quad,
\end{align}
%
where the expectations inside the integrals are given by
Equations~(\ref{eq:exp_y_sep}) and (\ref{eq:exp_yy_sep}), respectively, and
$\bar{\pmb{\theta}}_\star$ are all the stellar parameters \emph{except}
the inclination, i.e.,
%
\begin{align}
    \bar{\pmb{\theta}}_\star = \left(
    P \,\,\,
    u_1 \,\,\,
    u_2
    \right)^\top
    \quad,
\end{align}
%
where perhaps $P$ is fixed for each star and the limb darkening coefficients
are shared among all stars, as described above.
%
Thus, if we are able to perform the integrals in Equations~(\ref{eq:mu_marg})
and (\ref{eq:cov_marg}), we can \emph{dramatically} reduce the number of
parameters in our ensemble problem.
%
As we show in Appendix~\ref{sec:inclination}, if we assume that stellar
inclinations are distributed isotropically, these integrals
do in fact have closed-form solutions,
since they are (again) integrals over Wigner matrices.
%
We may then replace our expressions for the GP mean and covariance with
%
\begin{align}
    \label{eq:mu_marg}
    \pmb{\mu}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbf{f} \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big]
    \\
    \label{eq:cov_marg}
    \pmb{\Sigma}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
     & = \mathrm{E} \Big[ \mathbf{f} \, \mathbf{f}^\top \, \Big| \, \bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet \Big] - \pmb{\mu}(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet) \pmb{\mu}^\top(\bar{\pmb{\theta}}_\star, \pmb{\theta}_\bullet)
    \quad.
\end{align}
%
and use these to compute the marginal likelihood instead.

@@

\section{Computing the flux}
\label{sec:starry}

\subsection{Basic expression}
%
As we mentioned in \S\ref{sec:gp}, the flux $\mathbf{f}$ is a purely
linear function of the spherical harmonic coefficient vector $\mathbf{y}$:
%
\begin{align}
    \mathbf{f} = \mathbf{1} + \pmb{\mathcal{A}} \, \mathbf{y}
    \quad.
\end{align}
%
Even though this is derived in detail in \citet{Luger2019}, it is useful to
expand on the computation of the design matrix $\pmb{\mathcal{A}}$ in more
detail here. Let $\mathbf{a}_k^\top$ denote the $k^{th}$ row of $\pmb{\mathcal{A}}$,
such that
%
\begin{align}
    \label{eq:Arows}
    \pmb{\mathcal{A}}
     & =
    \setstackgap{L}{1.25\baselineskip}
    \fixTABwidth{T}
    \parenMatrixstack{
        \mathbf{a}_0^\top \\
        \mathbf{a}_1^\top \\
        \vdots            \\
        \mathbf{a}_{K-1}^\top
    }\quad.
\end{align}
%
The row vector $\mathbf{a}_k^\top$ encodes how the spherical harmonic
coefficient vector projects onto the $k^\mathrm{th}$ cadence in the flux timeseries, and
may be computed from
%
\begin{align}
    \label{eq:akT}
    \mathbf{a}_k^\top = \mathbf{r}^\top \,
    \mathbf{A_1} \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(-i\right) \,
    \mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P}t_k\right) \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(\frac{\pi}{2}\right) \,
    \mathbf{y}
    \quad.
\end{align}
%
To understand the expression above, let us proceed from right to left,
starting with the spherical harmonic vector $\mathbf{y}$, which we assume
describes the surface intensity of the star at time $t = 0$
in a frame where $\hat{\mathbf{x}}$
points to the right, $\hat{\mathbf{y}}$ points up, and $\hat{\mathbf{z}}$
points out of the page. The quantity $\mathbf{R}_{\hat{\mathbf{x}}}$
is a Wigner rotation matrix
(described in detail in \S\ref{sec:wigner}),
which in this case rotates the spherical harmonic representation
of the star by an angle $\nicefrac{\pi}{2}$ counter-clockwise about $\hat{\mathbf{x}}$
such that the north pole of the star points along $\hat{\mathbf{z}}$. In this
frame, we apply a second Wigner rotation matrix, $\mathbf{R}_{\hat{\mathbf{z}}}$,
to rotate the star about $\hat{\mathbf{z}}$ counter-clockwise (i.e., eastward) by an angle
$\nicefrac{2\pi t_k}{P}$, where $P$ is the rotation period and $t_k$ is the
time at cadence $t$.
Next, we rotate the star by a \emph{clockwise} angle of $i$ about $\hat{\mathbf{x}}$,
where $i$ is the stellar inclination ($i = 0$ corresponding to a pole-on view and
$i = \nicefrac{\pi}{2}$ corresponding to an edge-on view). With this last rotation,
we are now in the observer's frame%
\footnote{In principle, one last rotation could be performed about $\hat{\mathbf{z}}$
    to orient the projected disk of the star on the plane of the sky; however, the disk-integrated
    flux is independent of the rotation angle along the plane of the sky
    (which we refer to as the \emph{obliquity}), so this step is unnecessary.}.

Following \citet{Luger2019}, the next step is to project the representation of
the star into a more convenient basis for performing the integration over the stellar
disk. The change-of-basis matrix $\mathbf{A}_1$ \citep[c.f. Appendix~B in][]{Luger2019}
projects the stellar map into the \emph{polynomial basis}
\citep[Equation~7 in][]{Luger2019}, comprised of the sequence of monomials in Cartesian
coordinates
$\left( 1 \,\, x \,\, z \,\, y \,\, x^2 \,\, xz \,\, xy \,\, yz \,\, y^2 \cdots \right)$
where $z = \sqrt{1 - x^2 - y^2}$ on the surface of the unit sphere. We
can now compute the disk-integrated flux by integrating each of the terms in the basis
over the unit disk, which is straightforward in the polynomial basis; the
individual terms integrate to simple ratios of Gamma functions. These are then
assembled into the row vector $\mathbf{r}^\top$, given by Equation~(20) in
\citet{Luger2019}, which we dot into our expression (and add one) to obtain the
flux at the $k^\mathrm{th}$ cadence.

\subsection{With limb darkening}
%
\xxx{TODO.}

@@

As we mentioned earlier,
although we adopt $a$ and $b$ as the actual inputs to our Gaussian process
(i.e., as elements of the input vector $\pmb{\theta}_\bullet$), these
parameters are not themselves ``interpretable'', as they do not directly
correspond to physical
quantities of interest such as the mean and standard deviation of the spot
latitude distribution. However, over most of the parameter space in $a$ and $b$,
the spot latitude distribution is well approximated
by a bimodal Gaussian. In particular, there exists
a one-to-one relationship between $a$ and $b$ and the mean $\mu_\phi$
and standard deviation $\sigma_\phi$ of a normal approximation to the
distribution. Moreover, we find that
if we let $\mu_\phi$ be the \emph{mode} of the PDF and
$\sigma_\phi^2$ be a \emph{local} approximation to the variance of the PDF,
the relationship has a convenient closed form.

@@

In this section we will compute the first and second moments of the
inclination distribution (Equations~\ref{eq:einc} and \ref{eq:Einc}),
which allows us to compute the mean and covariance of the process
that describes the flux marginalized over all values of the inclination
(Equations~\ref{eq:mu_marg} and \ref{eq:cov_marg}).

\subsection{Probability density function}
%
Similar to the latitude integrals, the process of inclining a star relative to the
observer (see Equation~\ref{eq:akT}) involves
rotations by an angle $-i$ about $\hat{\mathbf{x}}$, which
may be accomplished by choosing
Euler angles $\alpha = \nicefrac{\pi}{2}$, $\beta = -i$, and
$\gamma = -\nicefrac{\pi}{2}$, such that
%
\begin{align}
    \mathbf{R}^l_{\hat{\mathbf{x}}}\left(-i\right)
     & =
    {\mathbf{U}^l}^{-1} \mathbf{D}^l_{\hat{\mathbf{x}}}\left(-i\right) \mathbf{U}^l
\end{align}
%
with
\begin{align}
    \mathbf{D}^l_{\hat{\mathbf{x}}}\left(-i\right)
     & =
    \mathbf{D}^l\left(\frac{\pi}{2}, -i, -\frac{\pi}{2}\right)
    \quad.
\end{align}
%
Since we expect an isotropic distribution of stellar rotation axes (absent
prior constraints on individual stars), the prior probability density for
the inclination $i$ is simply
%
\begin{align}
    p(i) = \sin i
\end{align}
%
for $i \in [0, \frac{\pi}{2}]$.

\subsection{First moment}
The expression for the first moment is
%
\begin{proof}{}
    \mathbf{e}_i
    & \equiv
    \int
    %\mathbf{R}_{\hat{\mathbf{x}}}(-i) \,
    \pmb{\mathcal{A}}(i, \bar{\pmb{\theta}}_\star) \,
    \mathbf{e}_y \,
    p(i) \,
    \mathrm{d}i
\end{proof}
%
where
%
\begin{proof}{}
    \mathbf{e}_y
    & \equiv
    \mathrm{E} \Big[ \mathbf{y} \, \Big| \, \pmb{\theta}_\bullet \Big]
\end{proof}
%
is the first moment of the GP in the spherical harmonic basis
(Equation~\ref{eq:exp_y}). Let us expand the expression above using
Equation~(\ref{eq:akT})

\begin{align}
    \label{eq:akT}
    \mathbf{a}_k^\top = \mathbf{r}^\top \,
    \mathbf{A_1} \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(-i\right) \,
    \mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P}t_k\right) \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(\frac{\pi}{2}\right) \,
    \mathbf{y}
    \quad.
\end{align}

\xxx{More soon.}
